---
title: 'Lab 2: R Basics Continued (Solutions)'
output: html_document
---
Today we'll use a collection of functions in the package called `tidyverse`. The `tidyverse` contains tons of extremely useful functions that make data cleaning and manipulation a breeze. 


## The Tidyverse

Let's use the `gapminder` dataset again. We can use the `filter()` function to select rows according to values of a column, and we can use `select()` to select only specific columns according to their name. If we want a specific set of rows, we can use `slice()`.

```{r, warning = F, message = F}
library(dplyr)
library(dslabs)
data(gapminder)
df <- gapminder
```

First, let's use `filter()`. Notice that we don't need to use `df$` in front of the names, since we supply `df` as the first argument to the function.

```{r}
# Filter
head(filter(df, country == 'United States'))
head(df[df$country == 'United States',])

# Notice that filter automatically throws out NA -- no need for which()!
head(filter(df, infant_mortality < 10))
head(df[df$infant_mortality < 10,])

# Multiple combinations
head(filter(df, infant_mortality < 10 | life_expectancy > 80))
```

1. How many rows have `infant_mortality > 20` and `life_expectancy > 70`?

```{r}
# If no NAs, these 2 things behave the same
# If NAs in one of these conditions, filter throws them out
# Base R returns rows where condition is TRUE and NA

df_problem1 <- filter(df, infant_mortality > 20, life_expectancy > 70)
nrow(df_problem1)

df_friend <- df[df$infant_mortality > 20 & df$life_expectancy > 70,]
nrow(df_friend)

### which() throws out NA
df_friend <- df[which(df$infant_mortality > 20 & df$life_expectancy > 70),]
nrow(df_friend)
```


Next we can use `select()` to get column names. `select()` doesn't require that we use quotes around names.

```{r}
# Select 3 columns (quotes work fine but not required)
head(select(df, country, year, infant_mortality))
head(select(df, 'country', 'year', 'infant_mortality'))

# Select everything except the region column
head(select(df, -region))
```

```{r}
### 1st row of df
slice(df, 1)

### rows 20-50 
slice(df, 20:50)
```


We can also order a data frame using the `arrange()` function.

```{r}
# Default is increasing
head(arrange(df, infant_mortality))
head(df[order(df$infant_mortality),]) # this is a little ugly

# Use desc() for decreasing
head(arrange(df, desc(infant_mortality)))
head(df[order(df$infant_mortality, decreasing = T),]) # even more ugly

# Sort by continent, then region
head(arrange(df, continent, region))
head(df[order(df$continent, df$region),]) # even more ugly


# First sort by continent [A-Z] then in descending order of infant mortality rate
df4 <- arrange(df, continent, desc(infant_mortality)) # much prettier/simpler!
```

2. Create a dataset that includes country, year, gdp and population for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by country, then year. Confirm that your code is doing what it's supposed to.

```{r}
# Subset Rows
df_complex <- filter(df, region == 'South-Eastern Asia', 
                     year >= 1965 & year <= 2010)

# Subset Columns
df_complex <- select(df_complex, country, year, gdp, population)

# Order Result
df_complex <- arrange(df_complex, country, year)
head(df_complex[order(df_complex$country, df_complex$year),])
```


### The pipe operator `%>%`

To generate the result above, you may have needed multiple steps or lines of code. We can use the pipe operator `%>%` to avoid using some of the intermediate steps and to cut down on the amount of code we need to type. This makes for much more intuitive and fun data wrangling.

For example, we can create the data frame mentioned above in an elegant, intuitive way:

```{r}
algeriagdp <- gapminder %>% 
  filter(country == 'Algeria') %>% 
  select(gdp, year)

# One way
x <- filter(gapminder, country == 'Algeria')
head(select(x, gdp, year))

# Another way (hard to read)
head(select(filter(gapminder, country == 'Algeria'), gdp, year))

```

3. Now, using the pipe operator, create a dataset that includes `country`, `year`, `gdp` and `population` for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by `country`, then `year`. Try to do it all in one command, starting from the `gapminder` dataset and using the pipe operator. Confirm that your code is doing what it's supposed to by printing out the result.

```{r}
# Subset rows
df_complex <- filter(df, region == 'South-Eastern Asia',
                     year >= 1965 & year <= 2010)

# Subset columns
df_complex <- select(df_complex, country, year, gdp, population)

# Order result
df_complex <- arrange(df_complex, country, year)

# Same result!
df_complex <- gapminder %>% 
  filter(region == 'South-Eastern Asia',
         year >= 1965 & year <= 2010) %>% 
  select(country, year, gdp, population) %>% 
  arrange(country, year)
```

### Creating variables

We can create variables using the function `mutate()`.

```{r}
gapminder <- mutate(gapminder, 'gdp_per_capita' = gdp/population)

# Same thing but using the pipe
gapminder <- gapminder %>% 
  mutate('gdp_per_capita' = gdp/population)
```

4. Create a variable with the approximate number of women in the population. Assume that 52% of the population are women.

```{r}
gapminder <- mutate(gapminder, 
                    'n_women' = 0.52 * population,
                    'n_men' = 0.48 * population)
```

5. Calculate an approximation of the number of children born each year based on the `fertility` column, which displays the expected number of children per woman over their reproductive life. For this, assume an average reproductive life of 25 years.

```{r}
gapminder <- gapminder %>% 
  mutate('children_per_year' = fertility * n_women/25)
```

--- 

You can find a list of our spells... ehhhm, functions [here](https://rstudio.github.io/cheatsheets/data-transformation.pdf).


And a list of other useful spellbooks (cheat sheets) [here](https://posit.co/resources/cheatsheets/). 

