---
title: "Combining tables"
output:
  html_document:
    toc: true
---

### TLDR: combining tables functions covered

* `inner_join(x, y)`: Keeps only the rows where keys match in both x and y. Adds new columns from y.

* `left_join(x, y)`: Keeps all rows from x, adds columns from y. Rows in x with no match in y get NA.

* `right_join(x, y)`: Keeps all rows from y, adds columns from x. Rows in y with no match in x get NA.

* `full_join(x, y)`: Keeps all rows from both x and y. Keys that don’t match appear with NA in the missing side.

* `semi_join(x, y)`: Keeps rows in x that have a matching key in y. No new columns are added.

* `anti_join(x, y)`: Keeps rows in x that do not have a matching key in y. No new columns are added.

**A handy way to remember:**

* `inner` = overlap only
* `left` = all of left + matches from right
* `right` = all of right + matches from left
* `full` = everything from both sides
* `semi` = filter x to matches in y
* `anti` = filter x to non-matches in y

**Binding:**

* `bind_rows()` stacks compatible tables vertically.

* `bind_cols()` glues tables side-by-side (same row count for each data frame).

**Set operations:**

* `intersect`, `setdiff`, and `setequal` treat rows as records; for vectors, they work on vector elements.


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
library(dslabs)
ds_theme_set()
```

The information we need for a given analysis may not be in just one table. Here we use a simple example to illustrate the general challenge of combining tables.

Suppose we want to explore the relationship between population size for US states, which we have in this table using the `murders` dataset from the `dslabs` package:

```{r}
data(murders)
head(murders)
```

and electoral votes (also from the `dslabs` package), which we have in this one:

```{r}
data(polls_us_election_2016)
head(results_us_election_2016)
```

Notice that just joining these two tables together will not work since the order of the states is not quite the same:

```{r}
identical(results_us_election_2016$state, murders$state)
```

The _join_ functions, described below, are designed to handle this challenge.

### Joins

The `join` functions in the `dplyr` package, which are based on SQL joins, make sure that the tables are combined so that matching rows are together.
The general idea is that one needs to identify one or more columns that will serve to match the two tables. Then a new table with the combined information is returned. Note what happens if we join the two tables above by `state` using `left_join`:

```{r}
tab <- left_join(murders, results_us_election_2016, by = "state")

tab %>% select(state, population, electoral_votes) %>% 
        head()
```

The data has been successfully joined and we can now, for example, make a plot to explore the relationship between population and electoral votes:

```{r}
tab %>% ggplot(aes(population/10^6, electoral_votes)) +
        geom_point() +
        scale_x_continuous(trans = "log2") +
        scale_y_continuous(trans = "log2") +
        geom_text_repel(aes(label = abb)) + 
        geom_smooth(method = "lm", se = FALSE) +
	      xlab("Population (in millions)") + 
	      ylab("Electoral Votes")
```

We see the relationship is close to linear with about 2 electoral votes for every million persons, but with smaller states getting a higher ratio.


In practice, it is not always the case that each row in one table has a matching row in the other. For this reason we have several different ways to join. To illustrate this challenge, take subsets of the matrices above:

```{r}
tab1 <- slice(murders, 1:6) %>% 
        select(state, population)
tab1
```

so that we no longer have the same states in the two tables:
```{r}
tab2 <- slice(results_us_election_2016, c(1:3, 5, 14, 44)) %>% 
        select(state, electoral_votes)
tab2
```

We will use these two tables as examples.

#### Left join

Suppose we want a table like `tab1` but adding electoral votes to whatever states we have available. For this we use left join with `tab1` as the first argument.

```{r}
left_join(tab1, tab2)
```

Note that `NA`s are added to the three states not appearing in `tab2`. Also note that this function, as well as all the other joins, can receive the first arguments through the pipe:

```{r}
tab1 %>% left_join(tab2)
```


#### Right join

If instead of a table like `tab1` we want one like `tab2` we can use `right_join`:

```{r}
tab1 %>% right_join(tab2)
```

Notice that now the NAs are in the column coming from `tab1`.

#### Inner join

If we want to keep only the rows that have information in both tables we use inner join. You can think of this an intersection:

```{r}
inner_join(tab1, tab2)
```

#### Full join

And if we want to keep all the rows, and fill the missing parts with NAs, we can use a full join. You can think of this as a union:

```{r}
full_join(tab1, tab2)
```

#### Semi join

The `semi_join` let's us keep the part of the first table for which we have information in the second. It does not add the columns of the second:

```{r}
semi_join(tab1, tab2)
```


#### Anti join

The function `anti_join` is the opposite of `semi_join`. It keeps the elements of the first table for which there is no information in the second:

```{r}
anti_join(tab1, tab2)
```

### Binding

Although we have yet to use it in this course, another common way in which datasets are combined is by _binding_ them. Unlike the join function, the binding functions do no try to match by a variable but rather just combine datasets. If the datasets don't match by the appropriate dimensions one obtains an error.

#### Columns

The `dplyr` function _bind_cols_ binds two objects by making them columns in a tibble. For example, if we quickly want to make a data frame consisting of numbers we can use.

```{r}
bind_cols(a = 1:3, b = 4:6)
```

This function requires that we assign names to the columns. Here we chose `a` and `b`. 

Note there is an R-base function `cbind` that performs the same function but creates objects other than tibbles. 

`bind_cols` can also bind data frames. For example, here we break up the `tab` data frame and then bind them back together:

```{r}
tab1 <- tab[, 1:3]
tab2 <- tab[, 4:6]
tab3 <- tab[, 7:9]
new_tab <- bind_cols(tab1, tab2, tab3)
head(new_tab)
```


#### Rows

The `bind_rows` is similar but binds rows instead of columns.

```{r}
tab1 <- tab[1:2,]
tab2 <- tab[3:4,]
bind_rows(tab1, tab2)
```

This is based on an R-base function `rbind`.

### Set Operators

Another set of commands useful for combing are the set operators. When applied to vectors, these behave as their names suggest. However, if the `tidyverse`, or more specifically, `dplyr` is loaded, these functions can be used on data frames as opposed to just on vectors.

#### Intersect

You can take intersections of vectors:

```{r}
intersect(1:10, 6:15)
```

```{r}
intersect(c("a","b","c"), c("b","c","d"))
```

But with `dplyr` loaded we can also do this for tables having the same column names:

```{r}
tab1 <- tab[1:5,]
tab2 <- tab[3:7,]
intersect(tab1, tab2)
```


#### Union

Similarly _union_ takes the union:

```{r}
union(1:10, 6:15)
```

```{r}
union(c("a","b","c"), c("b","c","d"))
```

But with `dplyr` loaded we can also do this for tables having the same column names:

```{r}
tab1 <- tab[1:5,]
tab2 <- tab[3:7,]
union(tab1, tab2)
```


#### Set differrence

The set difference between a first and second argument can be obtained with `setdiff`. Not unlike `instersect` and `union`, this function is not symmetric:


```{r}
setdiff(1:10, 6:15)
setdiff(6:15, 1:10)
```

As with the others above, we can apply it to data frames:
```{r}
tab1 <- tab[1:5,]
tab2 <- tab[3:7,]
setdiff(tab1, tab2)
```

#### `setequal`

Finally, the function `set_equal` tells us if two sets are the same, regardless of order. So

```{r}
setequal(1:5, 1:6)
```

but

```{r}
setequal(1:5, 5:1)
```

It also works when applied to data frames that are not equal regardless of order:

```{r}
setequal(tab1, tab2)
```

### Self-Assessment Questions

```{r}
# Setup
library(dslabs)
library(tidyverse)

data(murders)
data(gapminder)
data(research_funding_rates)
data(heights)
```

Set up data frames to use for most questions:

```{r}
library(dplyr)

# Table of employees
employees <- data.frame(
  id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Carol", "David")
)

# Table of departments
departments <- data.frame(
  id = c(2, 3, 5),
  dept = c("HR", "Finance", "IT")
)
```


1. Join employees and departments by id, only including ids that appear in both data frames. Which employees appear in the resulting data frame?

```{r}
# Your code here
```

2. Now join the data frames so that an employee's department is added if available.

```{r}
# Your code here
```

3. Now join the data frames so that an all departments are shown and an employee is added if available.

```{r}
# Your code here
```

4. Now combine both tables, keeping all ids.

```{r}
# Your code here
```

5. Now use a join function that results in a data frame with id and the name of an employee if their id appears in the departments data frame. In other words, a data frame that shows which employees have a matching department record, without having a column for department.

```{r}
# Your code here
```

6. Now use a join function that results in a data frame with id and the name of an employee if their id does not appear in the departments data frame. In other words, a data frame that shows which employees do not have a matching department record, without having a column for department.

```{r}
# Your code here
```

7. Use one of the functions we learned in this document to show that the resulting data frame in question 5 is distinct from the resulting data frame in question 6. 

```{r}
# Your code here
```


8. Using the `murders` dataset compute the murder rate per 100,000 and build two state sets: (A) states with murder rate > 3, (B) states in the South. Compute the intersection and set difference of A and B and explain what each means in words.

```{r}
# Your code here
```


### Self-Assessment Solutions

Set up data frames to use for most questions:

```{r}
library(dplyr)

# Table of employees
employees <- data.frame(
  id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Carol", "David")
)

# Table of departments
departments <- data.frame(
  id = c(2, 3, 5),
  dept = c("HR", "Finance", "IT")
)
```


1. Join employees and departments by id, only including ids that appear in both data frames. Which employees appear in the resulting data frame?

```{r}
employees %>%
  inner_join(departments, by = "id")
```

Bob and Carol appear in the resulting data frame. 

2. Now join the data frames so that an employee's department is added if available.

```{r}
# Solution #1
employees %>%
  left_join(departments, by = "id")

# Solution #2
departments %>%
  right_join(employees, by = "id")
```

3. Now join the data frames so that an all departments are shown and an employee is added if available.

```{r}
# Solution #1
employees %>%
  right_join(departments, by = "id")

# Solution #2
departments %>%
  left_join(employees, by = "id")
```

4. Now combine both tables, keeping all ids.

```{r}
employees %>%
  full_join(departments, by = "id")
```

5. Now use a join function that results in a data frame with id and the name of an employee if their id appears in the departments data frame. In other words, a data frame that shows which employees have a matching department record, without having a column for department.

```{r}
employees %>%
  semi_join(departments, by = "id")
```

6. Now use a join function that results in a data frame with id and the name of an employee if their id does not appear in the departments data frame. In other words, a data frame that shows which employees do not have a matching department record, without having a column for department.

```{r}
employees %>%
  anti_join(departments, by = "id")
```

7. Use one of the functions we learned in this document to show that the resulting data frame in question 5 is distinct from the resulting data frame in question 6. 

```{r}
# Save outputs of question 5 and 6 to use here
q5A <- employees %>%
  semi_join(departments, by = "id")
  
q6A <- employees %>%
  anti_join(departments, by = "id")

setequal(q5A, q6A)
```


8. Using the `murders` dataset compute the murder rate per 100,000 and build two state sets: (A) states with murder rate > 3, (B) states in the South. Compute the intersection and set difference of A and B and explain what each means in words.

```{r}
state_rates <- murders %>%
  mutate(rate = total / population * 100000)

A <- state_rates %>%
  filter(rate > 3) %>%
  .$state

B <- murders %>%
  filter(region == "South") %>%
  .$state

intersect(A, B)         # in both
setdiff(A, B)           # high rate but not in South
```

The intersection of A and B contains states that have a murder rate > 3 and are in the South. The set difference of A and B contains states with a murder rate > 3 but not in the South. 




