---
title: 'Lab 6: Machine Learning II'
output:
  html_document
---

In this lab we will first continue with breast cancer biopsy data from the University of Wisconsin. This dataset is available as `brca` in the `dslabs` package and is also part of the [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)). `brca` is a binary classification dataset for 569 malignant (cancer) and benign (not cancer) breast masses. 30 continuous features were computationally extracted from digital images of fine needle aspirate biopsy slides. 

For simplicity, let's focus on only the first two predictors, `radius_mean` and `texture_mean`. In the code chunk below, we construct a data frame called `data` with four variables: 

- `labels`: a factor with two levels denoting whether a mass is "Benign" or "Malignant". This will be useful for summarizing and plotting the data. 
- `y`: a numeric vector that encodes the outcomes as 0 for Benign and 1 for Malignant. This will be useful for model fitting. 
- `radius_mean`: the mean nucleus radius
- `texture_mean`: the mean nucleus texture

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(dslabs)
library(splitstackshape)
data(brca)

data = data.frame(labels = factor(ifelse(brca$y == "B", "Benign", "Malignant")), 
                  y = factor(ifelse(brca$y == "B", 0, 1)),
                  radius_mean = brca$x[,"radius_mean"], 
                  texture_mean = brca$x[,"texture_mean"])
```

Run the code below to create training and test sets. We will use a 70% training, 30% test set split. Note that we have an imbalance in outcome groups (212/569 = 37% of masses are malignant and 63% are not), so we need to take this into account when we split our data into training and test sets. Here we use the function `stratified` from the `splitstackshape` package to split the data. The stratified function samples the same percent of individuals from each class - in this case, `y`. Finally, `set.seed(219)` ensures reproducibility, so that we get the same random split every time we run the code. 

```{r, message=FALSE, warning=FALSE}
set.seed(219)

x           <- stratified(data, "y", 0.7, keep.rownames = TRUE)
train_set   <- x %>% dplyr::select(-rn) # Remove row numbers
train_index <- as.numeric(x$rn)
test_set    <- data[-train_index,]
test_set_y  <- as.factor(test_set$y)

head(train_set)
head(test_set)
```

## 1. k-Nearest Neighbors (KNN) for binary classification

### Question 1.1

Using 5-fold cross-validation, train a KNN model that predicts `y` based on `radius_mean` and `texture_mean`. Use `metric = "ROC"`. Then, use the `predict` function to get predicted probabilities for the test set, and convert the probabilities to predicted response labels (use a cutoff of 0.5). Be sure to standardize (preprocess) your data in the `train` function, and use values k = 3, 5, 7, ..., 31 for the `tuneGrid` argument. 

```{r}
set.seed(123)

cv5 <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

```{r}
k_grid <- data.frame(k = seq(3, 31, by = 2))

knn_cv <- train(
  labels ~ radius_mean + texture_mean,
  data = train_set,
  method = "knn",
  trControl = cv5,
  preProcess = c("center", "scale"),
  tuneGrid = k_grid,
  metric = "ROC"
)

knn_cv
```

```{r}
# Predictions on the test set
p_hat_knn_cv <- predict(knn_cv, newdata = test_set, type = "prob")[,2]

# Converting probabilities to 0 or 1
y_hat_knn_cv  <- factor(ifelse(p_hat_knn_cv >= 0.5, "Malignant", "Benign"),
                        levels = c("Benign","Malignant"))
```

### Question 1.2

Obtain the confusion matrix for the test set. Compare the performance of your KNN model with the Naive Bayes and logistic regression models from lab 5. 

```{r}
# Confusion matrix
cm_knn <- confusionMatrix(y_hat_knn_cv, test_set$labels, positive = "Malignant")
cm_knn
```

The KNN model has similar overall accuracy (0.883) to the Naive Bayes model (accuracy 0.87) and the logistic regression model (accuracy 0.906). The KNN model also has similar values for sensitivity and specificity with the 0.5 cutoff. For this 0.5 threshold and these predictors, the logistic regression model performs the best, the KNN model the second best, and the Naive Bayes model the worst. 


## 2. Regression and Decision Trees for Continuous Outcomes

In this part of the lab, we will predict infant birth weight in grams using the `birthwt` dataset, available in the `MASS` package. This dataset of 189 observations was  collected at the Baystate Medical Center in Springfield, MA during 1986. It includes the following variables: 

- `low`: indicator of birth weight less than 2.5 kg (0 = more than 2.5 kg, 1 = less than 2.5 kg).
- `age`: mother's age in years.
- `lwt`: mother's weight in pounds at last menstrual period.
- `race`: mother's race (1 = white, 2 = black, 3 = other).
- `smoke`: smoking status during pregnancy (0 = nonsmoker, 1 = smoker).
- `ptl`: number of previous premature labors.
- `ht`: history of hypertension (0 = no history, 1 = history of hypertension).
- `ui`: presence of uterine irritability (0 = no presence, 1 = presence of uterine irritability).
- `ftv`: number of physician visits during the first trimester.
- `bwt`: birth weight in grams.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(rpart)
library(MASS)

data(birthwt)
```

Since we are interested in predicting birth weight, we drop the `low` indicator from the dataset. We also convert `race`, `smoke`, `ht`, and `ui` to factors, because these variables are categorical and not numeric/continuous. 

Comment: Both the `dplyr` and `MASS` packages export a function called `select`, which is why the `dplyr::` is necessary in `dplyr::select`. This makes it clear to R that we want to use the `select` function from the `dplyr` package. 

```{r}
birthwt <- 
  birthwt %>% 
  dplyr::select(-low) %>%
  mutate(race = factor(race), 
         smoke = factor(smoke), 
         ht = factor(ht), 
         ui = factor(ui))
```

We use `createDataPartition` to split the birth weight data into training and test sets with 70% of the data in the training set and 30% in the test set. 

```{r}
set.seed(9)

# Create an index for training rows using caret
train_index <- createDataPartition(y = birthwt$bwt, p = 0.7, list = FALSE)

birthwt_train_set <- birthwt[train_index, ]
birthwt_test_set  <- birthwt[-train_index, ]

nrow(birthwt_train_set)
nrow(birthwt_test_set)
```


### Question 2.1

Based on the training set, make some plots to help you assess the relationship between our outcome of interest, `bwt`, and each of the other eight variables in the dataset. Which variables seem like good predictors of birth weight? 

**Solution:** 

For the continuous predictors (`age`, `lwt`, `ptl`, and `ftv`), we can make scatterplots to assess their relationships with infant birth weight.

```{r}
birthwt_train_set %>% 
  pivot_longer(cols = c('age', 'lwt', 'ptl', 'ftv'),
               names_to = 'predictor',
               values_to = 'value') %>% 
  ggplot(aes(x = value, y = bwt)) + 
  geom_point() + 
  facet_wrap(~ predictor, scales = 'free_x', 
             labeller = 
               as_labeller(c("age"="Mother's Age (years)", 
                             "lwt" = "Mother's Weight (lbs)", 
                             "ptl" = "Number of Premature Labors", 
                             "ftv" = "Number of 1st Trimester Visits"))) + 
  xlab("") + 
  ylab("Birth Weight (g)")
```

For the categorical predictors (`race`, `smoke`, `ht`, and `ui`), we can make boxplots. `ptl` (number of premature labors) and `ftv` (number of first trimester visits) are ordinal but not continuous, so they can also be visualized here as categorical predictors. You could potentially convert them to factors and have the models in Q2.2-2.4 treat them as categorical predictors, but we won't over-complicate things by doing that. 

Comment: The warning message "attributes are not identical across measure variables; they will be dropped" comes from gathering the predictor variables, which are factors that have different levels. It can be safely ignored in this case. 

```{r}
birthwt_train_set %>% 
  pivot_longer(cols = c('race', 'smoke', 'ht', 'ui'),
               names_to = 'predictor',
               values_to = 'value') %>% 
  ggplot(aes(x = value, y = bwt)) + 
  geom_boxplot() + 
  facet_wrap(~ predictor, scales = 'free_x', 
             labeller = 
               as_labeller(c("race"="Race", 
                             "smoke" = "Smoking Status", 
                             "ht" = "Hypertension", 
                             "ui" = "Uterine Irritability", 
                             "ptl" = "Number of Premature Labors", 
                             "ftv" = "Number of 1st Trimester Visits"))) + 
  xlab("") + 
  ylab("Birth Weight (g)")
```

It appears mothers who smoked during pregnancy or experienced uterine irritability may be more likely have babies with low birth weights. A few of the other predictors look like the may be associated with birth weight too, like mother's weight and race, but it's hard to say for certain.  

### Question 2.2

Fit a regression tree that predicts `bwt` using all of the other variables in the training data. You can use the `tree` function from the `tree` package. Make a plot that visualizes the tree.

**Solution:** 

The code for fitting a decision tree with `tree` follows a very similar syntax as the code for fitting a linear model with `lm`. When visualizing the tree, `pretty = 0` forces the text labels to use the factor level names stored in the categorical variables, instead of `a, b, ..., z`.

```{r}
library(rpart)
library(rpart.plot)

set.seed(1)

 # Fit a regression tree using all of the available predictors
fit_tree = rpart(bwt ~ ., birthwt_train_set)  

# Print a summary of the tree
summary(fit_tree)

# Plot the tree
rpart.plot(fit_tree, digits = 4)
```
The regression tree was constructed using `ui`, `race`, `lwt`, `smoke`, `age`, and `ftv`. 



### Question 2.3

For the model from Q2.2, calculate birth weight predictions for the observations in the test set. Calculate the RMSE and interpret it. 

**Solution:** 

You can get predictions for the model with a straightforward application of the `predict` function. You can also calculate MSE easily using the predictions and ground truth values. 

```{r}
# Use tree for prediction
preds <- predict(fit_tree, newdata = birthwt_test_set)

# Calculate RMSE
sqrt(mean((preds-birthwt_test_set$bwt)^2))
```

The MSE is equal to 630 grams. This tells us that the typical difference between our model’s predictions and the actual birthweight is 630 grams.

---

## 3. Decision Trees, Bagging, and Random Forests for Multi-Class Outcomes

In the final section of this lab, we will use gene expression data to classify tissue samples. The data can be loaded from the `dslabs` package by calling `data(tissue_gene_expression)`. `tissuesGeneExpression` is a list with two elements: 

- `x`: Numeric matrix with 189 rows and 500 columns. Each column contains gene expression measurements for a different gene. 
- `y`: Factor vector of length 189 that records tissue type labels (cerebellum, colon, endometrium, hippocampus, kidney, liver, or placenta) . 

The original data (accessible in the `tissuesGeneExpression` package) records gene expression for 22,215 genes. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(GGally)
library(tree)
library(randomForest)
library(dslabs)

data(tissue_gene_expression)
```

We will only use a random sample of 100 genes to predict tissue type. 

```{r}
set.seed(25)
tiss_ge <- 
  data.frame(y = tissue_gene_expression$y, 
             tissue_gene_expression$x[,sample(500, 100)])
```

As usual, we split the data into training and test sets, with 70% in the training set and 30% in the test set.

```{r}
set.seed(36)
tiss_ge_index_train <- createDataPartition(y = tiss_ge$y, 
                                           times = 1, p = 0.7, list = FALSE)

tiss_ge_train_set <- tiss_ge[tiss_ge_index_train, ]
tiss_ge_test_set  <- tiss_ge[-tiss_ge_index_train, ]

nrow(tiss_ge_train_set)
nrow(tiss_ge_test_set)
```


### Question 3.1

Below, you will find some plots and tables of the training set designed to help you develop some intuition for the data. Describe what you see. 

This is a frequency table for the tissue types in the training data. 

```{r}
table(tiss_ge_train_set$y)
```

It is difficult to make visualizations for all 100 genes in the dataset, so let's randomly sample six to focus on. 

```{r}
set.seed(49)
genes6 <- sample(names(tiss_ge)[-1], 6)
genes6
```

Here are histograms of the gene expression distributions of the six genes.  

```{r}
tiss_ge_train_set %>% 
  pivot_longer(cols = genes6,
               values_to = 'expression',
               names_to = 'gene') %>% 
  ggplot(aes(x = expression)) + 
  geom_histogram(color = "black") + 
  facet_wrap(~gene) + 
  xlab(NULL) + ylab(NULL)
```

The boxplots below plot gene expression against tissue type for the six genes. Note that setting `scales = 'free_y'` allows the y-axis to vary from plot to plot, so they are not on the same scale. 

```{r}
tiss_ge_train_set %>% 
  pivot_longer(cols = genes6,
               values_to = 'expression',
               names_to = 'gene') %>% 
  ggplot(aes(x = y, y = expression)) + 
  geom_boxplot() + 
  facet_wrap(~ gene, scales = 'free_y') + 
  xlab(NULL) + ylab(NULL) + 
  scale_x_discrete(labels = str_to_title(unique(tiss_ge_train_set$y))) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

The `ggcorr` function from the `GGally` package makes pretty correlation matrix plots. Each tiny square in this plot represents the correlation between a pair of genes (out of the entire set of 100 genes). Red indicates positive correlation (close to 1), blue indicates negative correlation (close to -1), and white indicates no correlation (close to 0). 

```{r}
ggcorr(tiss_ge_train_set[,-1], hjust = 1, size = 1.5, layout.exp = 10)
```

**Solution:** 

- From the frequency table, we can see that there is a non-uniform distribution of tissue types. Some tissues, like cerebellum and kidney, are much better represented in the data than others, like endometrium and placenta. 

- The histograms show that the gene expression measurements of the six randomly sampled genes have different distributions. Some are roughly symmetric, while others are skewed. COPB1 is highly expressed in most of the tissues, while RPE65 tends to be lowly expressed. 

- The boxplots suggest that individual genes are often expressed at different levels for different tissues. However, it doesn't really seem like any one of these six genes does a great job of distinguishing between all seven of the tissue types. 

- The correlation matrix plot (lots of red and blue squares) illustrates that many of the genes in this dataset are highly correlated with each other. So, we might not need all of the predictors to do a good job of classifying the tissues. 


### Question 3.2

Fit a bagging (bootstrap aggregation) model to the training data by running `randomForest` from the `randomForest` package with the `mtry` parameter set to the number of predictors (`mtry = 100`). Be sure to set a random seed so that your code is reproducible (the solutions use `set.seed(64)`). Get the predicted class labels for the test set data and report the test accuracy. 

**Solution:** 

Setting `mtry = 100` when we have 100 predictors is what makes this model a bagging model. 

```{r}
set.seed(64)
fit_bag <- randomForest(y ~ ., data = tiss_ge_train_set, mtry = 100)
```

By default, the `predict` function for `randomForest` models returns predicted class labels, but you can also explicitly specify `type = "response"`. 

```{r}
preds_fit_bag <- predict(fit_bag, newdata = tiss_ge_test_set)
```

You can use the `confusionMatrix` function to calculate diagnostic metrics for the test set predictions and pull out the accuracy from the `overall` slot. 

```{r}
confusionMatrix(preds_fit_bag, tiss_ge_test_set$y)$overall[1]
```

Alternatively, you can use the predicted and true test labels to calculate the accuracy by hand. 

```{r}
mean(preds_fit_bag == tiss_ge_test_set$y)
```

Either way, the test accuracy is 1, which is perfect accuracy. 


### Question 3.3

Now, build a random forest model with the `mtry` parameter set to the square root of the number of predictors. Also, set `importance = TRUE` so that the importance of the predictors is assessed. You will need the variable importance information for Q3.4. Be sure to set a random seed so that your code is reproducible (the solutions use `set.seed(81)`). Get the predicted class labels for the test set data and report the test accuracy. 

**Solution:** 

For classification problems, the default value for `mtry` is the square root of the number of predictors, but we can explicitly specify `mtry = 10`.  

```{r}
set.seed(81)
fit_rf <- randomForest(y ~ ., data = tiss_ge_train_set, 
                       mtry = 10, importance = TRUE)
```

As in Q3.2, you can calculate the accuracy directly or by using the `confusionMatrix` function. Here, the test accuracy is 1, meaning that all of the test set observations were classified perfectly by the model. We know from Q3.1 that the genes are highly correlated with each other. Random forests are known to decorrelate predictors, which explains the improvement over bagging. 

```{r}
preds_fit_rf <- predict(fit_rf, newdata = tiss_ge_test_set)

confusionMatrix(preds_fit_rf, tiss_ge_test_set$y)$overall[1]
mean(preds_fit_rf == tiss_ge_test_set$y)
```


### Question 3.4

Run the `importance` function on your random forest model from Q3.3 to extract variable importance measures for each of the tissue types. Find the five most important genes for classifying kidney tissues by ordering the Gini index measures.

**Solution:** 

The five most important genes for classifying kidney tissues are `COLGALT2`, `GPA33`, `AP3B2`, `USP32P2`, and `CELSR2`.  

```{r}
variable_importance <- importance(fit_rf)

variable_importance_kidney <- 
  data.frame(gene = rownames(variable_importance), 
             Gini = variable_importance[,"kidney"])

variable_importance_kidney %>% 
  arrange(desc(Gini)) %>% 
  head(5)
```


