---
title: "Coding question of the day"
output: html_document
---

### Sep 25, 2025

Using the gapminderdataset, calculate the mean fertility rate by region for the year 2015. Which 10 regions had the largest fertility rates that year? Assume there is no missing data. The following steps may be helpful.

1. Filter the data to only include the year 2015
2. Group by region
3. Calculate the average fertility rate for each region
4. Display the top 10 fertility rates (i.e., the 10 largest fertility rates)

```{r, message=FALSE}
library(dslabs)
library(dplyr)
data(gapminder)

gap_2015 <- gapminder %>%
  filter(year == 2015) %>%
  group_by(region) %>%
  summarise(avg_fertility = mean(fertility)) %>%
  arrange(desc(avg_fertility))

gap_2015
```

---

### Sep 30, 2025

Using the `gapminder` dataset, for the year 2000, add a column to the dataset called `high_life_exp` that is equal to 1 if the life expectancy is above 70 years of age and 0 otherwise. 

How many countries had a life expectancy above 70 years in 2000? What percentage of countries included in this dataset is this?

Hint: it might be helpful to use the `ifelse` function. 

```{r, message=FALSE}
library(dslabs)
library(dplyr)
data(gapminder)
```

```{r}
gapminder_2000 <- gapminder %>%
  filter(year == 2000) %>%
  mutate(high_life_exp = ifelse(life_expectancy > 70, 1, 0))

# Sum the number of countries with life expectancy above 70
sum(gapminder_2000$high_life_exp)

# Percentage calculation
dim(gapminder_2000) # Show the number of countries in the dataset to get denominator

100/185 # 54%

# Another way to do the calculation
sum(gapminder_2000$high_life_exp) / length(gapminder_2000$high_life_exp)
```

---

### Oct 2, 2025

Using the `gapminder` dataset and `ggplot2`, for the year 1986, plot `fertility` on the x-axis and `life_expectancy` on the y-axis. Color the points according to `continent`, and make the size of the points correspond to the population of the country. Update the x and y axis labels to make them more informative. Hint: the arguments color and size will be useful inside of ggplot. Note: you will most likely get ugly legends - don't worry about that right now.

```{r, message=FALSE}
library(dslabs)
library(dplyr)
library(ggplot2)
data(gapminder)
```

```{r}
gapminder %>% filter(year == 1986) %>%
  ggplot() +
  geom_point(aes(x = fertility, 
                 y = life_expectancy,
                 color = continent,
                 size = population)) +
  xlab("Fertility") +
  ylab("Life Expectancy")
```

---

### Oct 7, 2025

The `dslabs` package contains the `us_contagious_disease` dataset. This dataset includes weekly reported counts data (cases) for seven diseases from 1928 to 
2011 for all 50 US states. Use this dataset and `dplyr` tools to create an 
object called `data` that stores only the Measles data, includes a per 100,000 
people rate, and removes Alaska and Hawaii since they only became states in 
the late 1950s. Note there is a `weeks_reporting` column - take this into 
account when calculating the rate. There are 52 weeks in a year.

```{r}
library(dslabs)
library(dplyr)
data("us_contagious_diseases")

head(us_contagious_diseases)

```

```{r}
data <- us_contagious_diseases %>%
  filter(disease == "Measles",
         !state %in% c("Alaska", "Hawaii")) %>%
  mutate(rate = ((count/weeks_reporting)*52) / (population/100000))

head(data)
str(data)
glimpse(data)
```


```{r}
# Another way to do it

data <- us_contagious_diseases %>%
  filter(disease == "Measles",
         state != "Alaska",
         state != "Hawaii") %>%
  mutate(rate = ((count/weeks_reporting)*52) / (population/100000))

head(data)
str(data)
glimpse(data)
```

---

### Oct 9, 2025

Using the `gapminder` dataset, create a line graph of `life_expectancy` over time for the country Haiti. Make sure to include informative axis labels and a title. There is a substantial decrease in life expectancy one year - look up the reason for this if you don't know it already. Add text to your plot with `geom_text` indicating the event that happened that year that had such a drastic impact on life expectancy.

```{r}
library(dslabs)
library(dplyr)
data("gapminder")

```

