---
title: "Coding question of the day"
output: html_document
---

### Sep 25, 2025

Using the gapminderdataset, calculate the mean fertility rate by region for the year 2015. Which 10 regions had the largest fertility rates that year? Assume there is no missing data. The following steps may be helpful.

1. Filter the data to only include the year 2015
2. Group by region
3. Calculate the average fertility rate for each region
4. Display the top 10 fertility rates (i.e., the 10 largest fertility rates)

```{r, message=FALSE}
library(dslabs)
library(dplyr)
data(gapminder)

gap_2015 <- gapminder %>%
  filter(year == 2015) %>%
  group_by(region) %>%
  summarise(avg_fertility = mean(fertility)) %>%
  arrange(desc(avg_fertility))

gap_2015
```

---

### Sep 30, 2025

Using the `gapminder` dataset, for the year 2000, add a column to the dataset called `high_life_exp` that is equal to 1 if the life expectancy is above 70 years of age and 0 otherwise. 

How many countries had a life expectancy above 70 years in 2000? What percentage of countries included in this dataset is this?

Hint: it might be helpful to use the `ifelse` function. 

```{r, message=FALSE}
library(dslabs)
library(dplyr)
data(gapminder)
```

```{r}
gapminder_2000 <- gapminder %>%
  filter(year == 2000) %>%
  mutate(high_life_exp = ifelse(life_expectancy > 70, 1, 0))

# Sum the number of countries with life expectancy above 70
sum(gapminder_2000$high_life_exp)

# Percentage calculation
dim(gapminder_2000) # Show the number of countries in the dataset to get denominator

100/185 # 54%

# Another way to do the calculation
sum(gapminder_2000$high_life_exp) / length(gapminder_2000$high_life_exp)
```

---

### Oct 2, 2025

Using the `gapminder` dataset and `ggplot2`, for the year 1986, plot `fertility` on the x-axis and `life_expectancy` on the y-axis. Color the points according to `continent`, and make the size of the points correspond to the population of the country. Update the x and y axis labels to make them more informative. Hint: the arguments color and size will be useful inside of ggplot. Note: you will most likely get ugly legends - don't worry about that right now.

```{r, message=FALSE}
library(dslabs)
library(dplyr)
library(ggplot2)
data(gapminder)
```

```{r}
gapminder %>% filter(year == 1986) %>%
  ggplot() +
  geom_point(aes(x = fertility, 
                 y = life_expectancy,
                 color = continent,
                 size = population)) +
  xlab("Fertility") +
  ylab("Life Expectancy")
```

---

### Oct 7, 2025

The `dslabs` package contains the `us_contagious_disease` dataset. This dataset includes weekly reported counts data (cases) for seven diseases from 1928 to 
2011 for all 50 US states. Use this dataset and `dplyr` tools to create an 
object called `data` that stores only the Measles data, includes a per 100,000 
people rate, and removes Alaska and Hawaii since they only became states in 
the late 1950s. Note there is a `weeks_reporting` column - take this into 
account when calculating the rate. There are 52 weeks in a year.

```{r}
library(dslabs)
library(dplyr)
data("us_contagious_diseases")

head(us_contagious_diseases)

```

```{r}
data <- us_contagious_diseases %>%
  filter(disease == "Measles",
         !state %in% c("Alaska", "Hawaii")) %>%
  mutate(rate = ((count/weeks_reporting)*52) / (population/100000))

head(data)
str(data)
glimpse(data)
```


```{r}
# Another way to do it

data <- us_contagious_diseases %>%
  filter(disease == "Measles",
         state != "Alaska",
         state != "Hawaii") %>%
  mutate(rate = ((count/weeks_reporting)*52) / (population/100000))

head(data)
str(data)
glimpse(data)
```

---

### Oct 9, 2025

Using the `gapminder` dataset, create a line graph of `life_expectancy` over time for the country Haiti. Make sure to include informative axis labels and a title. There is a substantial decrease in life expectancy one year - look up the reason for this if you don't know it already. Add text to your plot with `geom_text` indicating the event that happened that year that had such a drastic impact on life expectancy.

```{r}
library(dslabs)
library(dplyr)
data("gapminder")

gapminder %>% filter(country == "Haiti") %>%
  ggplot(aes(year, life_expectancy)) +
  geom_line() +
  geom_vline(xintercept = 2010, lty = 2, color = "blue") +
  geom_text(aes(x = 2003, y = 34, label = "2010 Earthquake"), color = "blue") +
  xlab("Year") +
  ylab("Life Expectancy (Years)") +
  ggtitle("Life expectancy in Haiti")
```

---

### Oct 14, 2025

Using the `murders` dataset from the `dslabs` package and the `state.x77` dataset that is built in to R, make a scatterplot with the total number of gun murders (`total`) on the y-axis and land area in square miles (`Area`) on the x-axis. You will need to use the `left_join` function to join the `state.x77` dataset to the `murders` dataset using the `state` column. 

Bonus challenge: are there any NAs in the merged dataset? How are they created?

```{r}
# Load necessary libraries
library(dplyr)
library(dslabs)
library(ggplot2)
library(tidyr)

# Load the state.x77 dataset and convert to a data frame
data("state")
state_data <- as.data.frame(state.x77)

# Add the state names to the state.x77 dataset
state_data$state <- rownames(state_data)

head(state_data)

# Load the murders dataset from the dslabs package
data("murders")

# Use the left_join function to join the datasets by the state column
merged_data <- left_join(murders, state_data, by = "state")

# Plot area on x-axis and total number of murders on y-axis
merged_data %>% ggplot(aes(x = Area, y = total)) +
  geom_point() +
  labs(x = "Land area in square miles",
       y = "Total number of gun murders",
       title = "Number of gun murders vs land area across the US states")

# Bonus challenge
# Are there any NAs in the merged data set?
sum(is.na(merged_data)) # Yes, there are 8 NAs in the dataset

# Locate the row with NAs
merged_data[!complete.cases(merged_data),]

```

All of the NAs are in the District of Columbia row. All of the values from the state.x77 dataset are missing. This means that the state.x77 dataset did not contain a row for District of Columbia, which resulted in NA values when  joined to the murders dataset.

---

### Oct 16, 2025

```{r}
library(dslabs)
library(dplyr)
library(ggplot2)
data("murders")

murders %>% 
  mutate(rate = total / population * 100000,
         region = reorder(region, rate, FUN = median)) %>%
  ggplot(aes(region, rate, fill = region)) +
  geom_boxplot() +
  xlab("Region") +
  ylab("Murder rate per 100,000 persons") +
  ggtitle("2010 Gun Murder Rates Across US Regions") +
  guides(fill = "none") # removes legend since not needed here
```

---

### Oct 21, 2025

Using the gapminder dataset, make a line graph of life expectancy over time. Include one line per country, and another line with the global average life expectancy. Be sure to label the global average line and to make it stick out among the country-specific lines. Use the code below to calculate the crude global average for each year.

Hints:

1. Use alpha to help make the global average line stick out.
2. Use 2 geom_line layers.
3. Note that avg is a data frame.


```{r}
library(dslabs)
library(dplyr)
library(ggplot2)
data("gapminder")

avg <- gapminder %>%
  group_by(year) %>%
  summarize(global_avg = mean(life_expectancy))
```

---

### Oct 23, 2025

Using the `gapminder` dataset, create a categorical life expectancy variable and add it to the `gapminder` dataset. Call this variable `life_expectancy_category` and use the `case_when` function to create it. Life expectancy below 50 should be labeled "Low", life expectancy greater than or equal to 50 and less than 75 should be labeled "Medium", and life expectancy greater than or equal to 75 should be labeled "High".

Now, filter the dataset to only include observations from the year 1999, and create a bar chart of the 3 life expectancy categories. Reorder the bars so that they are in the order Low, Medium, High.


```{r}
library(dplyr)
library(dslabs)
library(ggplot2)

data("gapminder")
```













